---
title: "Q1"
output:
  html_document:
    df_print: paged
---


Install and load necessary packages
```{r}
options(repos = c(CRAN = "https://cloud.r-project.org/"))

install.packages("readxl")
install.packages("openxlsx")
install.packages("dplyr")
install.packages("tidyr")
install.packages("lubridate")

library(readxl)
library(openxlsx)
library(dplyr)
library(tidyr)
library(lubridate)
```


Path to the Excel file
```{r}
file_path <- "SAS_Data_Experienced.xlsx"
```


Read each sheet into a separate dataframe
```{r}
activity_df <- read_excel(file_path, sheet = "Activity")
diagnosis_df <- read_excel(file_path, sheet = "Diagnosis")
rx_df <- read_excel(file_path, sheet = "Rx")
```


Display the first few rows of each dataframe
```{r}
head(activity_df)
head(diagnosis_df)
head(rx_df)
```


Ensure that the date columns in diagnosis_df and rx_df are properly recognized as dates, convert 'Date' columns to Date type if not already
```{r}
diagnosis_df$Date <- as.Date(diagnosis_df$Date)
rx_df$`Rx Date` <- as.Date(rx_df$`Rx Date`)
```


---


### a) Export to Excel a list of patients that are continuously enrolled for at least 12 months at any point in time.


Reshape the activity data from wide to long format for easier sequential analysis and simplifies group operations
```{r}
activity_long <- activity_df %>%
  pivot_longer(
    cols = starts_with("M_"),
    names_to = "Month",
    values_to = "Enrolled"
  ) %>%
  arrange(Patient_ID, Month) %>%
  mutate(
    Year_Month = ymd(paste0(gsub("M_", "", Month), "01"))
  )
```


```{r}
activity_long
```


Helper function to check for 12 consecutive months of enrollment
```{r}
has_12_consecutive <- function(enrollments) {
  rle_result <- rle(enrollments)
  any(rle_result$values == 1 & rle_result$lengths >= 12)
}
```


Group by Patient_ID and check for 12 consecutive months
```{r}
patients_continuous <- activity_long %>%
  group_by(Patient_ID) %>%
  summarise(continuous_12_months = has_12_consecutive(Enrolled)) %>%
  filter(continuous_12_months == TRUE) %>%
  select(Patient_ID)
```


```{r}
patients_continuous
```


Export the resulting patient list to a new Excel file
```{r}
write.xlsx(patients_continuous, "patients_continuously_12_months.xlsx")
```


---


### b) Export to Excel a list of patients that are continuously enrolled for the most recent 12 months.


Extract the most recent 12 months for each patient
```{r}
recent_months <- activity_long %>%
  group_by(Patient_ID) %>%
  arrange(desc(Year_Month)) %>%
  slice_head(n = 12)
```


```{r}
recent_months
```


Check if all the recent 12 months show enrollment (1)
```{r}
recently_enrolled <- recent_months %>%
  group_by(Patient_ID) %>%
  summarise(All_Enrolled = all(Enrolled == 1)) %>%
  filter(All_Enrolled == TRUE) %>%
  select(Patient_ID)
```


```{r}
recently_enrolled
```


Export the resulting patient list to a new Excel file
```{r}
write.xlsx(recently_enrolled, "patients_most_recent_12_months.xlsx")
```


---


### c) Of patients in “b”, flag the patients that have received at least 2 ICD codes of any listed below (i.e. it could be 2 of the same kind or 1 of multiple different ones) in the most recent 12 months. i) 299.00, 399.88, 495.01


Read the list of patients from part (b) (most recent 12 months enrolled excel file)
```{r}
recently_enrolled <- read_excel("patients_most_recent_12_months.xlsx")
```


List of ICD codes of interest
```{r}
icd_codes_of_interest <- c("299.00", "399.88", "495.01")
```


Get the most recent date from the activity data
```{r}
most_recent_date <- max(activity_long$Year_Month)
one_year_ago <- most_recent_date %m-% months(11)
```


```{r}
print(most_recent_date)
print(one_year_ago)
```


Ensure that only the most recent patients patients' diagnoses are considered
```{r}
recent_diagnosis_df <- diagnosis_df %>%
  filter(
    Patient_ID %in% recently_enrolled$Patient_ID,
    Diag_Code %in% icd_codes_of_interest,
    Date >= one_year_ago & Date <= most_recent_date
  )
```


```{r}
recent_diagnosis_df
```


Filter the patients with at least 2 of the specified ICD codes in the most recent 12 months
```{r}
patients_with_at_least_2_icd <- recent_diagnosis_df %>%
  group_by(Patient_ID) %>%
  summarise(ICD_Count = n()) %>%
  filter(ICD_Count >= 2)
```


```{r}
patients_with_at_least_2_icd
```


---


### d) Of patients in “a”, flag the patients that have received at least 2 ICD codes of any listed below (i.e. it could be 2 of the same kind of 1 or multiple different ones) with a 12 month washout on the first claim (i.e. they are continuously enrolled for 12 months prior to their first claim) and 24 month look-forward period (i.e. they are continuously enrolled for 24 months after their first claim). These patients’ first ICD claim below will be referred to as “diagnosis claim” from here on out. i) 299.00, 399.88, 495.01


Read the list of patients from part (a) (patients continuously enrolled for at least 12 months)
```{r}
patients_continuous <- read_excel("patients_continuously_12_months.xlsx")
```


List of ICD codes of interest
```{r}
icd_codes_of_interest <- c("299.00", "399.88", "495.01")
```


Filter for patients from part (a) who received at least 2 ICD codes of interest
```{r}
diagnosis_with_2_icd <- diagnosis_df %>%
  filter(
    Patient_ID %in% patients_continuous$Patient_ID,
    Diag_Code %in% icd_codes_of_interest
  ) %>%
  group_by(Patient_ID) %>%
  summarise(ICD_Count = n(), First_Diagnosis_Date = min(Date)) %>%
  filter(ICD_Count >= 2)
```


```{r}
diagnosis_with_2_icd
```


Merge the diagnosis data with activity data to get enrollment history for these patients
```{r}
patient_diagnosis_activity <- activity_long %>%
  filter(Patient_ID %in% diagnosis_with_2_icd$Patient_ID) %>%
  left_join(diagnosis_with_2_icd, by = "Patient_ID") %>%
  mutate(
    Diagnosis_Washout_Period = interval(First_Diagnosis_Date %m-% months(12), First_Diagnosis_Date),
    Diagnosis_Look_Forward_Period = interval(First_Diagnosis_Date, First_Diagnosis_Date %m+% months(24))
  )
```


```{r}
patient_diagnosis_activity
```


Ensure that patients are continuously enrolled for 12 months before their first diagnosis claim 
```{r}
patients_with_washout <- patient_diagnosis_activity %>%
  group_by(Patient_ID) %>%
  filter(Year_Month <= First_Diagnosis_Date) %>%
  summarise(Enrolled_12_Months_Before = all(Enrolled == 1))
```


Ensure that patients are continuously enrolled for 24 months after their first diagnosis claim
```{r}
patients_with_look_forward <- patient_diagnosis_activity %>%
  group_by(Patient_ID) %>%
  filter(Year_Month >= First_Diagnosis_Date) %>%
  summarise(Enrolled_24_Months_After = all(Enrolled == 1))
```


Combine washout and look-forward results
```{r}
patients_with_2_icd_washout_and_look_forward <- patients_with_washout %>%
  inner_join(patients_with_look_forward, by = "Patient_ID") %>%
  inner_join(diagnosis_with_2_icd, by = "Patient_ID") %>%
  filter(Enrolled_12_Months_Before == TRUE & Enrolled_24_Months_After == TRUE)
```


```{r}
patients_with_2_icd_washout_and_look_forward
```


---


### e) Of patients in “d”, find the number of patients that were treated with the drugs (below) after their diagnosis (do not look further than 24 months after diagnosis). i) Abilify ii) Seroquel


List of drugs of interest
```{r}
drugs_of_interest <- c("Abilify", "Seroquel")
```


Filter the Rx data for patients from part (d) who were prescribed the drugs of interest
```{r}
patients_with_treatment <- rx_df %>%
  filter(
    Patient_ID %in% patients_with_2_icd_washout_and_look_forward$Patient_ID,
    Drug %in% drugs_of_interest
  )
```


```{r}
patients_with_treatment
```


Join with the diagnosis data to get the first diagnosis date for each patient
```{r}
patients_with_treatment <- patients_with_treatment %>%
  left_join(diagnosis_with_2_icd, by = "Patient_ID")
```


```{r}
patients_with_treatment
```


Filter the prescriptions that were made within 24 months after the diagnosis
```{r}
patients_treated_within_24_months <- patients_with_treatment %>%
  filter(`Rx Date` >= First_Diagnosis_Date & `Rx Date` <= (First_Diagnosis_Date %m+% months(24)))
```


```{r}
patients_treated_within_24_months
```


Count the number of patients who were prescribed Abilify or Seroquel within the 24-month period after diagnosis
```{r}
individual_counts <- patients_treated_within_24_months %>%
  group_by(Drug) %>%
  distinct(Patient_ID) %>%
  summarise(Num_Patients = n())

total_patients_treated <- patients_treated_within_24_months %>%
  distinct(Patient_ID) %>%
  count()

print(individual_counts)
print(total_patients_treated)
```


---


### f) Of patients in “e”, report the patients that have a 12+ month washout on their first treatment claim post diagnosis for i) Abilify ii) Seroquel


List of drugs of interest
```{r}
drugs_of_interest <- c("Abilify", "Seroquel")
```


Filter the Rx data for patients from part (e)
```{r}
patients_with_treatment <- rx_df %>%
  filter(
    Patient_ID %in% patients_treated_within_24_months$Patient_ID,
    Drug %in% drugs_of_interest
  )
```


```{r}
patients_with_treatment
```


Join with diagnosis data to get the first diagnosis date for each patient
```{r}
patients_with_treatment <- patients_with_treatment %>%
  left_join(diagnosis_with_2_icd, by = "Patient_ID")
```


```{r}
patients_with_treatment
```


Identify the first treatment claim post-diagnosis
```{r}
patients_first_treatment <- patients_with_treatment %>%
  filter(`Rx Date` >= First_Diagnosis_Date & `Rx Date` <= (First_Diagnosis_Date %m+% months(24))) %>%
  group_by(Patient_ID, Drug) %>%
  summarise(First_Treatment_Date = min(`Rx Date`))
```


```{r}
patients_first_treatment
```


Check for 12+ month washout period prior to the first treatment
```{r}
patients_with_prior_treatment <- patients_first_treatment %>%
  left_join(rx_df, by = c("Patient_ID", "Drug")) %>%
  filter(`Rx Date` < First_Treatment_Date & `Rx Date` >= (First_Treatment_Date %m-% months(12)))
```


```{r}
patients_with_prior_treatment
```


Exclude patients who had any prior treatment in the 12 months before first treatment
```{r}
patients_with_washout_treatment <- patients_first_treatment %>%
  filter(!(Patient_ID %in% patients_with_prior_treatment$Patient_ID))
```


```{r}
patients_with_washout_treatment
```


---


### g) Of patients in “f”, report the mean/median days of supply they received in the first 12 months after their initial treatment claim for the drugs below (do not look further than 24 months after diagnosis). i) Abilify ii) Seroquel


Join the filtered washout patients with the prescription data (rx_df)
```{r}
patients_days_supply <- patients_with_washout_treatment %>%
  left_join(rx_df, by = c("Patient_ID", "Drug")) %>%
  left_join(diagnosis_with_2_icd, by = "Patient_ID") %>%
  filter(
    `Rx Date` >= First_Treatment_Date & 
    `Rx Date` <= (First_Treatment_Date %m+% months(12)) &
    `Rx Date` <= (First_Diagnosis_Date %m+% months(24))
  ) %>%
  group_by(Patient_ID, Drug) %>%
  summarise(
    Mean_Days_Supply = mean(Days_Supply, na.rm = TRUE),
    Median_Days_Supply = median(Days_Supply, na.rm = TRUE)
  )
```


```{r}
patients_days_supply
```

